Persistence in functional topology and a correction to a theorem of Morse

### 2021-07-16 persistence.tex

Usual constructions that work in the category of vector spaces over the field $\F$ can be transferred to the category of persistence modules by applying them pointwise.
For example, the kernel and cokernel of a morphism, as well as the direct sum of persistence modules are well-defined.
Persistence modules that are \emph{indecomposable}, i.e., those that have only trivial direct sum decompositions, play an important role in persistence theory.
A rich family of indecomposable persistence modules is given by \emph{interval modules}, which for an interval $I \subseteq \R$ are defined by
\begin{equation} \label{e:interval module}
    C(I)_t =
    \begin{cases}
        \mathbb{F} & \text{if } t \in I, \\
        0          & \text{otherwise},
    \end{cases}
    \qquad
    \qquad
    C(I)_{s, t} =
    \begin{cases}
        \operatorname{id}_{\mathbb{F}} & \text{if } s, t \in I,\\
        0 & \text{otherwise}.
    \end{cases}    
\end{equation}
These indecomposable interval modules can be used as building blocks for \emph{barcode modules}, which are direct sums of interval modules.
Given a barcode module $\bigoplus_{\lambda \in \Lambda} C(I_{\lambda})$, the associated multiset of intervals $\{I_{\lambda}\}_{\lambda \in \Lambda}$ is known as its \textit{barcode}.
By a version of the Krull--Remak--Schmidt--Azumaya Theorem \cite{MR37832} (see also \cite[Theorem 2.7]{Chazal.2016a} for a specialization to barcode modules), two isomorphic barcode modules have the same barcodes up to a choice of the index set $\Lambda$.
Thus, if a persistence module $M$ is provided with an isomorphism to a barcode module, referred to as a \emph{barcode decomposition}, the associated barcode is a complete isomorphism invariant of $M$.
So understanding which persistence modules admit barcode decompositions is key.
The most commonly used existence result for barcode decompositions is due to Crawley-Boevey's \cite{Crawley-Boevey.2015}.
It guarantees the existence of a barcode decomposition for any \emph{pointwise finite dimensional} ($\PFD$) persistence module, which is a persistence modules $M$ such that $M_t$ is a finite dimensional vector space for all $t \in \R$.
Unfortunately, the $\PFD$ condition is too restrictive for many purposes.
In particular, it is unsuited for the the applications of Morse and Tompkins in minimal surfaces theory.
An appropriate weakening of $\PFD$ for more general settings is the notion of \emph{q-tame}.
A persistence module $M$ is q-tame if the rank of the map $M_{s,t} \colon M_s \to M_t$ is finite dimensional for all $s < t$  \cite{Chazal.2016a}.
As exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$, not every q-tame persistence module admits a barcode decomposition in the above sense.
Yet, there are multiple ways of working around this to still get discrete invariants in the spirit of barcodes \cite{Chazal.2016a, Chazal.2016b, schmahl2020structure}.
We briefly recall the approach from \citet{Chazal.2016b}.

@@

Instead of talking about the barcode of the radical of a q-tame persistence module, it will be more convenient for our treatment of Morse inequalities to talk about the \emph{(undecorated) persistence diagram} associated to a barcode $\{I_{\lambda}\}_{\lambda \in \Lambda}$, which is the multiset defined by the \emph{multiplicity function} $\mathfrak{m} \colon \multiplicityDomain \to \mathbb{N}$ that associates to an element in
\[
\multiplicityDomain =
\big\{ (p,q) \mid p \in \R \cup \{-\infty\}, \ q \in \R \cup \{+\infty\}, \ p < q \big\}
\]
the cardinality of the set $\{ \alpha \in A \mid \inf I_{\alpha} = p,\ \sup I_{\alpha} = q\}$.

>>>>

Usual constructions that work in the category of vector spaces over the field $\F$ can be transferred to the category of persistence modules by applying them pointwise.
For example, the kernel and cokernel of a morphism, as well as the direct sum of persistence modules are well-defined.
Persistence modules that are \emph{indecomposable}, i.e., those that have only trivial direct sum decompositions, play an important role in persistence theory.
A rich family of indecomposable persistence modules is given by \emph{interval modules}, which for an interval $I \subseteq \R$ are defined by
\begin{equation} \label{e:interval module}
    C(I)_t =
    \begin{cases}
        \mathbb{F} & \text{if } t \in I, \\
        0          & \text{otherwise},
    \end{cases}
    \qquad
    \qquad
    C(I)_{s, t} =
    \begin{cases}
        \operatorname{id}_{\mathbb{F}} & \text{if } s, t \in I,\\
        0 & \text{otherwise}.
    \end{cases}    
\end{equation}
These indecomposable interval modules can be used as building blocks for \emph{barcode modules}, which are direct sums of interval modules.
Given a barcode module $\bigoplus_{\lambda \in \Lambda} C(I_{\lambda})$, the associated multiset of intervals $\{I_{\lambda}\}_{\lambda \in \Lambda}$ is known as its \textit{barcode}.
By a version of the Krull--Remak--Schmidt--Azumaya Theorem \cite{MR37832} (see also \cite[Theorem 2.7]{Chazal.2016a} for a specialization to barcode modules), two isomorphic barcode modules have the same barcodes up to a choice of the index set $\Lambda$.
Thus, if a persistence module $M$ is provided with an isomorphism to a barcode module, referred to as a \emph{barcode decomposition}, the associated barcode is a complete isomorphism invariant of $M$.
So understanding which persistence modules admit barcode decompositions is key.
The most commonly used existence result for barcode decompositions is due to Crawley-Boevey's \cite{Crawley-Boevey.2015}.
It guarantees the existence of a barcode decomposition for any \emph{pointwise finite dimensional} ($\PFD$) persistence module, which is a persistence modules $M$ such that $M_t$ is a finite dimensional vector space for all $t \in \R$.
Unfortunately, the $\PFD$ condition is too restrictive for many purposes.
In particular, it is unsuited for the the applications of Morse and Tompkins in minimal surfaces theory.
An appropriate weakening of $\PFD$ for more general settings is the notion of \emph{q-tame}.
A persistence module $M$ is q-tame if the rank of the map $M_{s,t} \colon M_s \to M_t$ is finite for all $s < t$ \cite{Chazal.2016a}.
As exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$, not every q-tame persistence module admits a barcode decomposition in the above sense.
Yet, there are multiple ways of working around this to still get discrete invariants in the spirit of barcodes \cite{Chazal.2016a, Chazal.2016b, schmahl2020structure}.
We briefly recall the approach from \citet{Chazal.2016b}.

@@

Instead of talking about the barcode of the radical of a q-tame persistence module, it will be more convenient for our treatment of Morse inequalities to talk about the \emph{(undecorated) persistence diagram} associated to a barcode $\{I_{\lambda}\}_{\lambda \in \Lambda}$, which is the multiset defined by the \emph{multiplicity function} $\mathfrak{m} \colon \multiplicityDomain \to \mathbb{N}$ that associates to an element in
\begin{equation*}
\multiplicityDomain =
\big\{ (p,q) \mid p \in \R \cup \{-\infty\}, \ q \in \R \cup \{+\infty\}, \ p < q \big\}
\end{equation*}
the cardinality of the set $\{ \alpha \in A \mid \inf I_{\alpha} = p,\ \sup I_{\alpha} = q\}$.
